Precise spiking motifs in neurobiological and neuromorphic data

### 2022-11-14 manuscript.tex
Figure Haimerl 2019

From another viewpoint, there is a substantial literature in neurobiology indicating that brain dynamics often organize into stereotyped sequences such as synfire chains~\citep{ikegaya_synfire_2004}, packets~\citep{luczak_sequential_2007} or hippocampal sequences~\citep{pastalkova_internally_2008}. %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Explictly decoding spiking motifs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%We have previously seen that neurobiologists have described stereotyped sequences of neural activation, notably in the hippocampus~\citep{pastalkova_internally_2008,villette_internally_2015,malvache_awake_2016}, and that neurons can be activated in the same order across days~\citep{haimerl_internal_2019}. 
Going further, researchers have found that  precise repetitions of spontaneous patterns of synaptic inputs in neocortical neurons, \textit{in vivo} and \textit{in vitro}. These patterns repeat after minutes, maintaining millisecond accuracy. Indeed,~\citet{ikegaya_synfire_2004} demonstrated that in cortical activity, one can find a repetition of several motifs in spike activity (duration around 1s +/- .5 s, some events in motifs are of similar size but sometimes absent). These sequences can be specific of a particular layer or column, can be synchronized with network activity oscillation, and can involve several cells. They also demonstrated that these sequences can form super sequences, so-called \emph{cortical songs}. It consists of the assembly of several sequences which repeat in a specific order with a compressed timing. \emph{In vivo} spontaneous activity also reveals repeating sequences: about 3000 sequences, each involving 3-10 cells out (of about 900), and lasting up to 3 seconds. Sequences have specific topographic structures, in some cases involving only a particular layer or a vertical column of cells or cells located in a cluster, and are associated with a structured spatial organization of the neurons that formed them. %In cortical songs, there is a ``compressing timing'' which may be taken into account by a similar mechanism as maxpooling in CNNs for space, but in time. Or there may be a mechanism for controlling the replay speed (pulvinar, \ldots{} , ?)
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=.6\linewidth]{figures/Ikegaya2004zse0150424620001.jpeg}%pcbi.1006283.g001.PNG_L.png}
\caption{Fig. 1. from~\citep{ikegaya_synfire_2004} repeated motifs of spontaneous synaptic activity \textit{in vitro} and \textit{in vivo}. (A) Repeated motifs of intracellular activity from layer 5 pyramidal neurons in slices. Panels show segments (red) of the same voltage-clamp recording from the same cell, repeating seconds or minutes after the initial occurrence (blue). Arrows indicate timings of repeated PSCs. (i) Upper trace: low--temporal resolution display of spontaneous activity of a neuron. Lower traces: higher resolution display of the repeated motif at indicated regions of the trace (a to c). (ii) Example of a longer motif. (B) Three repetitions of a motif. The top traces show the motifs superimposed on each other (blue, green, and red), the middle traces show these same traces individually, and the bottom traces show temporally magnified regions of the motifs (a to c). (C) Repeated sequences of intracellular current-clamp recordings \textit{in vivo}. Two (i) and three (ii) repetitions of motifs are shown. Shuffle tests were performed on traces (i), a to c, yielding significantly fewer repeats (fig.~S2, P  0.02). In (i), the blue trace is shifted --2.75 mV; in (ii), the blue trace is shifted --1.58 mV, and the green +0.79 mV.}\label{fig:Ikegaya2004}
\end{figure}
%----------------------------%

Additional studies detail the role of such precise spike timing in downstream information transfer and coding~\citep{villette_internally_2015,branco_dendritic_2010,luczak_packet-based_2015}. This is for instance relevant in sensory pathways in vision~\citep{meister_concerted_1995}, audition~\citep{decharms_primary_1996}, olfaction~\citep{wehr_odour_1996} or touch~\citep{johansson_first_2004}.  In particular, stereotyped sequences of neural activation have been described in the adult hippocampus and related to its function in mental travel in time and space~\citep{buzsaki_space_2018}. These sequences can be internally generated~\citep{pastalkova_internally_2008,villette_internally_2015} and may be formed by the chained activation of orthogonal assemblies, themselves organized as sequence packets~\citep{malvache_awake_2016}. In that protocol, hippocampal sequences are formed by the ordered activation of smaller sequence motifs. They are stereotyped and robust, since neurons can be activated in the same order across days (see Figure~\ref{fig:haimerl} from~\citep{haimerl_internal_2019}). As a consequence, hippocampal sequences may rely on an internally hardwired structure and form the functional building blocks for encoding, storing and retrieving experience.

%
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=\linewidth]{figures/haimerl2019.jpg}
\caption{An example of a polychronous group seen in cortical slices. In this study by~\citep{haimerl_internal_2019}, an analysis of the raster plot shows repetitions of precise spiking motifs with a time scale of the order of seconds. (A) Calcium fluorescence (heatmap) of CA1 neurons participating to run sequences in consecutive imaging sessions. Cells have been selected and ordered with respect to their activity in the first imaging session (Top). The black line represents the speed of the mouse.}\label{fig:haimerl}
\end{figure}
%----------------------------%

>>>>

From another viewpoint, there is a substantial literature in neurobiology indicating that brain dynamics often organize into stereotyped sequences such as synfire chains~\citep{ikegaya_synfire_2004}, packets~\citep{luczak_sequential_2007} or hippocampal sequences~\citep{pastalkova_internally_2008}. %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Explictly decoding spiking motifs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%We have previously seen that neurobiologists have described stereotyped sequences of neural activation, notably in the hippocampus~\citep{pastalkova_internally_2008,villette_internally_2015,malvache_awake_2016}, and that neurons can be activated in the same order across days~\citep{haimerl_internal_2019}. 
Going further, researchers have found that  precise repetitions of spontaneous patterns of synaptic inputs in neocortical neurons, \textit{in vivo} and \textit{in vitro}. These patterns repeat after minutes, maintaining millisecond accuracy. Indeed,~\citet{ikegaya_synfire_2004} demonstrated that in cortical activity, one can find a repetition of several motifs in spike activity (duration around 1s +/- .5 s, some events in motifs are of similar size but sometimes absent). These sequences can be specific of a particular layer or column, can be synchronized with network activity oscillation, and can involve several cells. They also demonstrated that these sequences can form super sequences, so-called \emph{cortical songs}. It consists of the assembly of several sequences which repeat in a specific order with a compressed timing. \emph{In vivo} spontaneous activity also reveals repeating sequences: about 3000 sequences, each involving 3-10 cells out (of about 900), and lasting up to 3 seconds. Sequences have specific topographic structures, in some cases involving only a particular layer or a vertical column of cells or cells located in a cluster, and are associated with a structured spatial organization of the neurons that formed them. %In cortical songs, there is a ``compressing timing'' which may be taken into account by a similar mechanism as maxpooling in CNNs for space, but in time. Or there may be a mechanism for controlling the replay speed (pulvinar, \ldots{} , ?)
% %----------------------------%
% \begin{figure}
% \centering
% \includegraphics[width=.6\linewidth]{figures/Ikegaya2004zse0150424620001.jpeg}%pcbi.1006283.g001.PNG_L.png}
% \caption{Fig. 1. from~\citep{ikegaya_synfire_2004} repeated motifs of spontaneous synaptic activity \textit{in vitro} and \textit{in vivo}. (A) Repeated motifs of intracellular activity from layer 5 pyramidal neurons in slices. Panels show segments (red) of the same voltage-clamp recording from the same cell, repeating seconds or minutes after the initial occurrence (blue). Arrows indicate timings of repeated PSCs. (i) Upper trace: low--temporal resolution display of spontaneous activity of a neuron. Lower traces: higher resolution display of the repeated motif at indicated regions of the trace (a to c). (ii) Example of a longer motif. (B) Three repetitions of a motif. The top traces show the motifs superimposed on each other (blue, green, and red), the middle traces show these same traces individually, and the bottom traces show temporally magnified regions of the motifs (a to c). (C) Repeated sequences of intracellular current-clamp recordings \textit{in vivo}. Two (i) and three (ii) repetitions of motifs are shown. Shuffle tests were performed on traces (i), a to c, yielding significantly fewer repeats (fig.~S2, P  0.02). In (i), the blue trace is shifted --2.75 mV; in (ii), the blue trace is shifted --1.58 mV, and the green +0.79 mV.}\label{fig:Ikegaya2004}
% \end{figure}
% %----------------------------%

Additional studies detail the role of such precise spike timing in downstream information transfer and coding~\citep{villette_internally_2015,branco_dendritic_2010,luczak_packet-based_2015}. This is for instance relevant in sensory pathways in vision~\citep{meister_concerted_1995}, audition~\citep{decharms_primary_1996}, olfaction~\citep{wehr_odour_1996} or touch~\citep{johansson_first_2004}.  In particular, stereotyped sequences of neural activation have been described in the adult hippocampus and related to its function in mental travel in time and space~\citep{buzsaki_space_2018}. These sequences can be internally generated~\citep{pastalkova_internally_2008,villette_internally_2015} and may be formed by the chained activation of orthogonal assemblies, themselves organized as sequence packets~\citep{malvache_awake_2016}. In that protocol, hippocampal sequences are formed by the ordered activation of smaller sequence motifs. They are stereotyped and robust, since neurons can be activated in the same order across days (see Figure~\ref{fig:haimerl} from~\citep{haimerl_internal_2019}). As a consequence, hippocampal sequences may rely on an internally hardwired structure and form the functional building blocks for encoding, storing and retrieving experience.
%
%----------------------------%
\begin{figure}
\centering
\includegraphics[width=\linewidth]{figures/haimerl2019.jpg}
\caption{An example of a polychronous group seen in cortical slices. In this study by~\citep{haimerl_internal_2019}, an analysis of the raster plot shows repetitions of precise spiking motifs with a time scale of the order of seconds. (A) Calcium fluorescence (heatmap) of CA1 neurons participating to run sequences in consecutive imaging sessions. Cells have been selected and ordered with respect to their activity in the first imaging session (Top). The black line represents the speed of the mouse (modified from Figure 1 from~\citep{haimerl_internal_2019} under the CC-BY \href{https://www.pnas.org/doi/full/10.1073/pnas.1718518116}{PNAS License}.).}\label{fig:haimerl}
\end{figure}
%----------------------------%
